A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-15 tex/codatadp.tex
Refactor lemma.

For any $\mu$-term $t$ of the form $\mu x. C( \vec u )$, we may recursively construct a normal $\mu$-term $\nf{t}$
(which we will call the \emph{normal form} of $t$)
that is a renaming of $v$ by replacing all of the bisimilar subterms of $t$ with $x$,
%and by replacing all of the nonnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For instance, $\nf{(\mu x. \const{C}( \mu y. \const{C}( x ) ))} = \mu x. \const{C}( x )$.

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how wrong-applied selector terms and variables in $\Ec$ are interpreted.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
We compute the \emph{completion} of $\Val$ for normal model $\M$, which we will write as $\Val^\ast$,
which we construct by assigning values from $\M$ to unassigned variables in the domain of $\Val$.

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x} : \tau$ of $\Ec$,
\item
$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one (call the value in its domain $c$).
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign $c$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Cyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.

>>>>

For any $\mu$-term $t$ of the form $\mu x. C( \vec u )$, we may recursively construct a normal $\mu$-term $\nf{t}$
(which we will call the \emph{normal form} of $t$)
that is a renaming of $t$ by replacing all of the bisimilar subterms of $t$ with $x$,
%and by replacing all of the nonnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For instance, $\nf{(\mu x. \const{C}( \mu y. \const{C}( x ) ))} = \mu x. \const{C}( x )$.

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how $\M$ interprets wrong-applied selector terms and variables.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
We compute the \emph{completion} of $\Val$ for normal model $\M$, which we will write as $\Val^\ast$,
which we construct by assigning values from $\M$ to unassigned variables in the domain of $\Val$.

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x} : \tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu v_{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one (call the value in its domain $c$).
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign $c$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Cyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.
